
\matlab is a dynamic scientific language used by scientists, engineers and
students worldwide.  Although \matlab is very suitable for rapid prototyping
and development,  \matlab users
often want to convert their final \matlab programs to a static language such as {\sc
FORTRAN}, to integrate them into already existing programs of that language,
to leverage the performance of powerful static compilers, or to 
ease the distribution of executables.

This thesis presents an extensible object-oriented toolkit to help
facilitate the generation of static programs from dynamic \matlab
programs.  Our open source toolkit, called the \matlab Tamer, targets
a large subset of \matlab. Given information about the entry point of
the program, the \matlab Tamer builds a complete callgraph, transforms
every function into a reduced intermediate representation, and
provides typing information to aid the generation of static code.

In order to provide this functionality, we need to handle a large
number of \matlab builtin functions. Part of the Tamer framework is
the builtin framework, an extensible toolkit which provides a
principled approach to handle a large number of builtin functions.  To
build the callgraph, we provide an interprocedural analysis framework,
which can be used to implement full-program analyses.  Using this
interprocedural framework, we have developed value analysis, an extensible
interprocedural analysis to estimate \matlab types, which helps
discover the call edges needed to build the call graph.

In order to make the static analyses even possible, we disallow a
small number of \matlab constructs and features, but attempt to
support as large a subset of \matlab as possible.  Thus, by both
slightly restricting \matlab, and by providing a framework with
powerful analyses and simplifying transformations, we can
``Tame \matlab''.




